---
title: "Peak_alignment_and_matching"
author: "Pietro Franceschi"
date: "30/08/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r message=FALSE}
## load the library
library(xcms)
## this library allows the "interactive" plots
library(plotly)
```


## Introduction
Peak detection will result in a set of "peak lists" which are not immediately comparable. Several phenomena are responsible for that

* mass accuracy is never infinite, so even the same ion will be never detected exactly at the same m/z
* chromaography is never perfectly reproducible
* interesting samples are never identical ...

To construct a reasonably good data matrix we have then to match the indivisual peaklists into a consensus list of "features".
Mass shifts are usually small and easier to control - at the end of the story we implement a sort of "binning" -, to compensate for chromatographic shifts a specific alignment step (also colled retention time correction) is required.  



